---
title: "Part 2: Multiple Imputation"
author: "Hannah Han"
output:
  github_document
---

# 1. Preparation

## Import pacakges {.smaller}

```{r}
setwd("/Users/hannah/Documents/GitHub/CodingClub_workshop/2.0-Missing_Data")
# install.packages(c("GGally", "mice", "tidyverse", "finalfit"))
library(tidyverse) # for data wraggling
library(finalfit) # great package to handle missing data
library(GGally) # for visualizing missing pattern
library(mice) # for multiple imputation
```

## Load demo data (if needed need) {.smaller}

```{r}
# select a subset for multiple imputation exercise
load( here::here("data", "processed", "cchs2015_demonstration.rdata"))
  
# select a subset of multiple imputation practice
MD <- cchs2015_demonstration %>%
  select(participantid, age, sex, bmi,
         energy, protein, cho, fat)

# view MD subset
str(MD)
summary(MD)

```

-   missing values are represented with NA
-   variables with missing: bmi

# 2. Our question

Mock question of interest: **is energy intake associated with with fat intake?**

-   Dependent: energy intake
-   Main predictor: fat intake
-   Covariates of interest: age, sex, bmi
-   Auxiliary variables:
    -   Variables that are either correlated with missing variable (r \> 0.4) or believed to be associated with missingness. *Use your knowledge!*
    -   To increase power and/or help make the assumption of MAR more plausible
    -   protein, cho

# 3. Explore pattern of missingness

```{r}
explanatory = c("protein", "age", "sex", "bmi")
dependent = "energy"

# look for patterns of missingness
MD %>%
 missing_pattern(dependent, explanatory)
```

-   there are two pattern of missingness

```{r}
# explore pattern of missingness
MD %>%
  missing_pairs(dependent, explanatory, position = "fill")
```

-   pattern of missingness
    -   *blue*: observed values
    -   *grey*: missing values
    -   the distribution of observed and missing BMI can be visually compared

```{r}
# compare not missing vs. missing 
explatory_compare = c("protein", "age", "sex", "energy")
dependent_compare = "bmi"
MD %>%
  missing_compare(dependent_compare, explatory_compare) %>%
  knitr::kable(row.names=FALSE, align = c("l", "l", "r", "r", "r"))
```

-   difference in estimated protein and energy intake
    -   does not satisfy MCAR, more likely to be MAR
    -   would not recommend complete case analysis

# 4. deciding how to handle the data: multiple imputation

## 4.1 imputation phase {.smaller}

-   *Dependent*: energy intake
-   *Main predictor*: fat intake
-   *Covariates of interest*: age, sex, bmi
-   *Auxiliary variables*: protein, cho

```{r}
# impute missing values for all variables using mice package
# in our case, we would just need to include all variables in MD
# number of imputation: 10
multiple_imputation = mice(
  MD,
  seed = 123,
  m = 10,
  print = FALSE) 
```

**note**: mice has many prediction methods including Predictive Mean Matching, Chained Equations, Random Forests, etc. We are using the default Predictive Mean Matching

## 4.2 analysis and pooling phase {.smaller}

```{r}
model_fit <-
  with(multiple_imputation, lm(energy ~ fat + age + sex + bmi))

base::summary(mice::pool(model_fit)) %>% 
  tibble::as_tibble() |>
   knitr::kable(
     digits=2
   )

```

# 5. potential things to think about

-   choose your auxiliary variables carefully! make sure to include all variables that you **think** are associated with or predict missingness!
-   number of imputation: Recommendation varies. Rule of thumb:
    -   5-20 imputations: low fractions of missingness
    -   as many as 50 or more if needed: high fraction of missingness
    -   bottom line: coefficient eventually stabilize after certain number of imputation. Can run different time of m to assess stability
-   about transformed variables: recommend to just treat them as another variable and back transform if necessary
-   **not covered but need to know**: there are ways to assess imputation quality!

# 6. Reference

-   Multiple Imputation in SAS. UCLA: Statistical Consulting Group. From https://stats.oarc.ucla.edu/sas/seminars/multiple-imputation-in-sas/mi_new_1/ (accessed April 27, 2023)
-   Batra, Neale, et al. The Epidemiologist R Handbook. Chapter 20 Missing Data. 2021.
-   Enders, C. K. (2022). Applied Missing Data Analysis. Guilford Publications. https://books.google.ca/books?id=b294EAAAQBAJ
