---
title: "R Notebook"
author: "Hannah Han"
output:
  github_document:
    df_print: paged
---

# 1. Preparation
## Import pacakges {.smaller}
```{r}
# install.packages(c("GGally", "mice", "tidyverse", "finalfit"))
library(tidyverse) # for data wraggling
library(finalfit) # great package to handle missing data
library(GGally) # for visualizing missing pattern
library(mice) # for multiple imputation
```
## Load demo data (if needed need) {.smaller}
```{r}
load("~/Documents/GitHub/CodingClub_workshop/data/processed/cchs2015_demonstration.rdata")

# select a subset for multiple imputation exercise
MD <- cchs2015_demonstration %>%
  select(participantid, age, sex, bmi,
         energy, protein, cho, fat)

# view MD subset
str(MD)
summary(MD)

```
* missing values are represented with NA
* variables with missing: bmi

# 2. Our question
__Mock question of interest__: is energy intake associated with with fat intake?
- _Dependent_: energy intake
- _Main predictor_: fat intake 
- _Covariates of interest_: age, sex, bmi
- _Auxiliary variables_: 
    + Variables that are either correlated with missing variable (r > 0.4) or believed to be associated with missingness. _Use your knowledge!_
    + To increase power and/or help make the assumption of MAR more plausible
- protein, cho

# 3. Explore pattern of missingness
```{r}
explanatory = c("protein", "age", "sex", "bmi")
dependent = "energy"

# look for patterns of missingness
MD %>%
 missing_pattern(dependent, explanatory)
```
- there are two pattern of missingness

```{r}
# explore pattern of missingness
MD %>%
  missing_pairs(dependent, explanatory, position = "fill")
```
- pattern of missingness
    + _blue_: observed values
    + _grey_: missing values
    + the distribution of observed and missing BMI can be visually compared
    
```{r}
# compare not missing vs. missing 
explatory_compare = c("protein", "age", "sex", "energy")
dependent_compare = "bmi"
MD %>%
  missing_compare(dependent_compare, explatory_compare) %>%
  knitr::kable(row.names=FALSE, align = c("l", "l", "r", "r", "r"))
```

- difference in estimated protein and energy intake
    + does not satisfy MCAR, more likely to be MAR
    + would not recommend complete case analysis

# 4. deciding how to handle the data: multiple imputation
## 4.1 imputation phase {.smaller}
- _Dependent_: energy intake
- _Main predictor_: fat intake 
- _Covariates of interest_: age, sex, bmi
- _Auxiliary variables_: protein, cho
```{r}
# impute missing values for all variables using mice package
# in our case, we would just need to include all variables in MD
# number of imputation: 10
multiple_imputation = mice(
  MD,
  seed = 123,
  m = 10,
  print = FALSE) 
```
__note__: mice has many prediction methods including Predictive Mean Matching, Chained Equations, Random Forests, etc. We are using the default Predictive Mean Matching

## 4.2 analysis and pooling phase {.smaller}
```{r}
model_fit <- with(multiple_imputation, lm(energy ~ fat + age + sex + bmi))

base::summary(mice::pool(model_fit))
```

# 5. potential things to think about
- choose your auxiliary variables carefully! make sure to include all variables that you __think__ are associated with or predict missingness!
- number of imputation: Recommendation varies. Rule of thumb:
    + 5-20 imputations: low fractions of missingness
    + as many as 50 or more if needed: high fraction of missingness
    + bottom line: coefficient eventually stabilize after certain number of imputation. Can run different time of m to assess stability
- about transformed variables: recommend to just treat them as another variable and back transform if necessary
- __not covered but need to know__: there are ways to assess imputation quality!

# 6. Reference
- Multiple Imputation in SAS. UCLA: Statistical Consulting Group. From https://stats.oarc.ucla.edu/sas/seminars/multiple-imputation-in-sas/mi_new_1/ (accessed April 27, 2023)
- Batra, Neale, et al. The Epidemiologist R Handbook. Chapter 20 Missing Data. 2021.
- Enders, C. K. (2022). Applied Missing Data Analysis. Guilford Publications. https://books.google.ca/books?id=b294EAAAQBAJ 






